---
title: "Getting started with PGA-cmdr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with PGA-cmdr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## Overview

**PGA-cmdr** is an interactive Shiny application for exploratory and
descriptive population genetics analyses from individual-based multilocus
datasets (microsatellites, codominant markers). All computations run
locally in R/C++ (Rcpp, OpenMP) -- no data is sent to any server.

It covers the full analytical workflow:

1. Data import, with automatic loci-range and encoding detection
2. Allele frequencies per locus and population
3. General statistics -- Ho, Hs, Ht, FIT/FIS/FST (Weir & Cockerham 1984),
   plus optional Fst-max, Fst' (Meirmans), GST/GST'' (Nei/Hedrick)
4. Null allele frequency estimation (FreeNA EM algorithm, ENA correction)
5. Local panmixia -- FIS per locus and/or population (Weir & Cockerham 1984)
6. Global panmixia -- FIT across all populations
7. Population subdivision -- FST per locus and globally, plus a G-test
8. Genetic diversities -- HS, HT (Nei & Chesser 1983), with resampling over
   individuals, sub-samples, or loci
9. Linkage disequilibrium -- pairwise G-test with permutation p-values
10. Isolation by distance -- Rousset's (1997) regression and a Mantel test

Every module follows the same pattern: set parameters, click **Run**, and a
`.zip` (or a single `.txt` for one-table modules) downloads automatically
with everything you need -- there is nothing to preview or export
separately first. The exact output file name(s) are always shown just
above the Run button before you click it.

## Launch the application

```{r launch}
library(shinypopgen)
run_app()
```

The application opens in your default browser.

## Data format

PGA-cmdr accepts CSV or tabulation-delimited TXT files. The separator
(comma, semicolon, or tab) is detected automatically, and the file loads
as soon as it is chosen -- no extra click needed. The file must contain a
header row, one *population* column, and one or more *genetic marker*
columns; other columns (individual ID, latitude, longitude, etc.) are
simply ignored. Two genotype encodings are supported, but a single file
must use only one of them throughout.

### Single-column format

Each locus occupies one column; alleles are separated by a character
(`/`, `-`, `_`, etc.):

```
Population  B12      C07      D12
Pop1        192/194  145/192  92/112
Pop1        200/200  179/179  108/112
Pop2        145/145  92/92    92/100
```

### Paired-column format

Each locus occupies two *adjacent* columns, one per allele. The first
allele column carries the locus name (e.g. `B12`); the second must
immediately follow it and be named `B12_1` (or `B12.1`):

```
Population  B12  B12_1  C07  C07_1
Pop1        192  194    145  192
Pop1        200  200    179  179
Pop2        145  145     92   92
```

Missing genotypes are coded as `0` (or `0/0`) by default and are
customisable on import.

## Setting the loci range on import

Once the file is loaded, a numbered **Column reference** table appears at
the bottom of the Import Data page, listing every column name next to its
position -- use it to check exactly which number corresponds to which
column.

Fill in:

1. **Population / Latitude / Longitude** -- pick the matching column from
   each dropdown (Latitude/Longitude are optional, needed only for the
   IBD module's automatic geographic-distance calculation).
2. **Column number of the first locus** -- where your markers start.
3. **Number of loci** -- the *actual number of genetic markers* (not raw
   columns). As soon as the first locus column is set, this field is
   **auto-suggested** from how many complete loci fit between that column
   and the end of the file, correctly accounting for single- or
   paired-column format. Adjust it down if some trailing columns aren't
   loci.

The app shows the resulting column range live (e.g. "6 loci x 2
columns/locus = columns 9 to 20") before you click **Apply**.

## Worked example: *Boophilus* cattle ticks

The package ships a demonstration dataset from Thierry de Meeus' teaching
material (`inst/extdata/default_dataset.csv`): adult *Boophilus* ticks
sampled from cattle across New Caledonia, genotyped at 6 microsatellite
loci (B12, C07, D12, D10, A12, C03) across 8 sampling localities.

The steps below follow the app's recommended workflow (see the in-app
**Help** tab for the full reference list).

### 1. Import the data

Click **"Load Default Data"** in the sidebar of the *Import Data* tab, or
follow the manual steps above for your own file. The application imports
the data via DuckDB, builds the internal genotype table (individual x
locus, packed integers), and is ready for every other module immediately.

### 2. Allele frequencies

Set an output name on the *Allele Freq* tab (or leave it auto-filled) and
click **Run**. A single `.zip` downloads with per-locus, per-population
allele frequencies plus a population-size-weighted global column -- every
population and marker is included automatically.

### 3. General statistics

The *General Stats* tab reports, per locus: Ho, Hs, Ht, and FIT/FIS/FST
(Weir & Cockerham 1984), plus any of the optional indices you tick
(Fst-max, Fst', GST, GST''). The same Run also produces a per-population
summary, per-locus detail for every population, and per-allele
F-statistics -- all in one zip.

### 4. Null alleles

If a locus's Ho/Hs ratio looks unusually low, check the *Null Alleles*
tab: null allele frequencies are estimated per locus x sub-sample by the
FreeNA EM algorithm (Chapuis & Estoup 2007), with both raw and
ENA-corrected FST/DCSE reported, and a full pairwise table.

### 5. Local panmixia (FIS)

The *Local Panmixia* tab reports FIS (Weir & Cockerham 1984) both **by
locus** and **by population** in the same output -- no need to choose one
view, both are always computed. Each comes with a bootstrap confidence
interval and a permutation p-value. Sub-sample or individual counts below
5 are reported as `NA` rather than an unreliable estimate.

### 6. Global panmixia (FIT)

The *Global Panmixia* tab reports the multilocus FIT across all
populations, with the same bootstrap/permutation approach.

### 7. Subdivision (FST)

The *Subdivision* tab estimates FST (Weir & Cockerham 1984) per locus and
overall, with bootstrap CI (over loci and over sub-samples) and a
permutation p-value; a separate genotype-reassignment G-test (as in
FSTAT, Goudet et al. 1996) is also available, reporting two one-sided
p-values (heterozygote deficit / excess) matching FSTAT's own convention.

### 8. Genetic diversities (HS, HT)

The *Diversities* tab reports HS (within-population) and HT (total) gene
diversity -- using the **unbiased small-sample estimator of Nei & Chesser
(1983)**, not the simpler biased form -- per locus and as multilocus
estimates, with confidence intervals from resampling over individuals,
sub-samples, or loci.

### 9. Linkage disequilibrium

The *LD* tab tests every locus pair, per population and combined ("All"),
using a permutation-based log-likelihood-ratio (G) test. One locus's
genotypes are permuted among individuals within each population
(10,000 permutations by default) and compared to the observed G statistic.

```{r ld-note, eval=FALSE}
# G = 2 * sum(observed * log(observed / expected)), under independence
```

### 10. Isolation by distance

The *IBD* tab offers two independent analyses:

- **Rousset's (1997) regression** of linearised genetic distance
  (FST/(1-FST)) against geographic distance -- this module **reuses the
  pairwise table already computed by the Null Alleles module** rather
  than recomputing it, so run Null Alleles first (or upload an external
  pairwise file). Geographic distance is computed automatically from
  Latitude/Longitude set at import.
- **A Mantel permutation test** (Pearson, Spearman, and Rousset 1D/2D
  statistics), independent of the regression above.

## Statistical methods

All F-statistics follow **Weir & Cockerham (1984)**:

> Weir, B.S. & Cockerham, C.C. (1984). Estimating F-statistics for the
> analysis of population structure. *Evolution*, **38**, 1358--1370.

Gene diversity (HS, HT) follows the unbiased small-sample estimator of
**Nei & Chesser (1983)**:

> Nei, M. & Chesser, R.K. (1983). Estimation of fixation indices and gene
> diversities. *Annals of Human Genetics*, **47**, 253--259.

Null allele frequencies and the ENA correction follow **Chapuis & Estoup
(2007)** (the FreeNA method):

> Chapuis, M.P. & Estoup, A. (2007). Microsatellite null alleles and
> estimation of population differentiation. *Molecular Biology and
> Evolution*, **24**, 621--631.

Isolation by distance follows **Rousset (1997)**:

> Rousset, F. (1997). Genetic differentiation and estimation of gene flow
> from F-statistics under isolation by distance. *Genetics*, **145**,
> 1219--1228.

The genotype-reassignment G-test for subdivision follows **FSTAT**:

> Goudet, J., Raymond, M., de Meeus, T. & Rousset, F. (1996). Testing
> differentiation in diploid populations. *Genetics*, **144**, 1933--1940.

Fst-max and Fst' follow **Meirmans (2006)** and **Hedrick (2005)**; GST''
follows **Meirmans & Hedrick (2011)**:

> Meirmans, P.G. & Hedrick, P.W. (2011). Assessing population structure:
> FST and related measures. *Molecular Ecology Resources*, **11**, 5--18.

Permutation p-values use the formula `p = (b + 1) / (m + 1)`, where `b` is
the number of permuted statistics at least as extreme as the observed
value and `m` is the number of permutations, following:

> Phipson, B. & Smyth, G.K. (2010). Permutation p-values should never be
> zero: calculating exact p-values when permutations are randomly drawn.
> *Statistical Applications in Genetics and Molecular Biology*, **9**, 39.

Heavy computations (bootstrap, permutation loops) are parallelised with
OpenMP via C++17 (Rcpp).

## A note on cross-module dependencies

Most modules are fully self-contained: they read the imported dataset and
compute everything themselves. The one exception is **Isolation by
Distance**, whose Rousset regression reuses the Null Alleles module's
pairwise table instead of recomputing it. If you go directly to IBD
without having run Null Alleles first, the app will tell you so and ask
you to run it (or upload an external pairwise file instead). Keep this in
mind if you change your imported data or parameters between the two --
re-run Null Alleles before trusting IBD's regression output.

## Session info

```{r session}
sessionInfo()
```
